Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-04-15 2020/RSD_GPU/RSD2020.tex
Add comment for the last table

Problems of multidimensional multiextremal optimization and numerical methods for their solution are considered. The general assumption is made about the function being optimized: it satisfies the Lipschitz condition with an a priori unknown constant. Many approaches to solving problems of this class are based on reducing the dimension of the problem; i.e., addressing a multidimensional problem by solving a family of problems with lower dimension. In this work, an adaptive dimensionality reduction scheme is investigated, and its implementation using graphic accelerators is proposed. Numerical experiments on several hundred test problems were carried out, and they confirmed acceleration in the developed GPU version of the algorithm.

@@

However, the potential for graphics accelerators to solve global optimization problems has not yet been fully realized. Using the GPUs, they basically parallelize nature-inspired optimization algorithms, which are somehow based on the idea of random search (see, for example, \cite{Ferreiro2013,Garcia2014,Langdon2011}). By virtue of their stochastic nature, algorithms of this type guarantee convergence to the global minimum only in the sense of probability, which differentiates them unfavorably from deterministic methods.

With regard to many deterministic algorithms of Lipschitzian global optimization with guaranteed convergence, parallel variants have been proposed \cite{Evtushenko2009,He2008,Paulavicius2011}. However, these versions of algorithms are parallelized on the CPU using shared and/or distributed memory; presently, no GPU implementations have been made. For example, \cite{Paulavicius2011} describes parallelization of an algorithm based on the ideas of the branch and boundary method using MPI and OpenMP.

Within the framework of this research, we consider the problems of capturing the optimum, which are characterized by a lengthy period for calculating the values of objective function in comparison with the time needed for processing them. For example, objective function can be specified using systems of linear algebraic equations, systems of ordinary differential equations, etc. Currently, graphics accelerators can be used to solve problems of this type. Moreover, an accelerator can solve several such problems at once \cite{Kindratenko2014}; i.e., using the GPU, one can calculate multiple function values simultaneously.

Thus, calculating the optimization criterion can be implemented on the GPU, and the role of the optimization algorithm (running on the CPU) consists in the effective selection of points for conducting parallel trials. This scheme of working with the accelerator is fully consistent with the work of the parallel global search algorithm developed at the Lobachevsky State University of Nizhni Novgorod and presented in a series of publications \cite{Barkalov2014,Barkalov2020,Barkalov2019,Grishagin2016,Grishagin2015,Grishagin2018,globalizerSystem}.

@@

In relation to global optimization methods, an operation that can be efficiently implemented on GPU is the parallel calculation of many values of the objective function at once. Naturally, this requires implementing a procedure for calculating the value of a function on GPU. Data transfers from CPU to GPU will be minimal: one need only transfer the coordinates of the trial points to the GPU, and get back the function values at these points. Functions that determine the processing of trial results in accordance with the algorithm, and require work with large amounts of accumulated search information, can be effectively implemented on CPU.

@@

\ref{fig:Parallel_Computing} shows the general scheme of organizing calculations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\varphi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using only the CPU. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

@@

The numerical experiments were carried out using the Lomonosov supercomputer (Lomonosov Moscow State University). Each supercomputer node included two quad-core processors, Intel Xeon X5570 2.93GHz and 12 Gb RAM. When building the Globalizer system for running on the Lomonosov supercomputer, the GCC 4.3.0 compiler, CUDA 6.5 and Intel MPI 2017 were used.

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time span for calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series).

@@

Let us now evaluate the acceleration achieved using parallel GSA using an adaptive dimensionality reduction scheme based on the number $p$ of cores used.
Table \ref{table:average_MMAS_CPU} shows the speedup of the algorithm that combines multiple evolvents and an adaptive scheme for solving a series of problems on CPU, compared to the sequential launch of the GSA. Two evolvents and, accordingly, two processes were used; each process used $p$ threads, and calculations were performed on a single cluster node.
% Table \ref{table:average_AS_CPU} shows the speedup obtained when solving a series of problems on the CPU, compared to GSA.

@@

\begin{table}
                \caption{Results on GPU}
                \label{table:average_AS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    5,06  & 3,93  &   & 1,19 	 & 5,41  &   & 2,2	    & 1,92	 &   & 0,47	& 2,21 \\
                               $p=256$ &    19,91 & 14,95 &   & 11,9 	 & 23,36 &   & 8,28	& 6,91	     &   & 3,4	    & 6,1 \\
							   $p=1024$&    15,62 & 52,91 &   & 22,12	 & 55,22 &   & 5,21	& 19,95      &   & 2,18	& 4,82 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on GPU}
                \label{table:average_MMAS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,32	 & 6,83	 &   & 8,19	     & 10,66  &   & 2,44	 & 1,63	 &   & 1,33	 & 1,58 \\
                               $p=256$ &    38,96	 & 31,52 &   & 25,43	 & 48,23  &   & 8,01	 & 7,04	 &   & 2,31	 & 3,42 \\
							   $p=1024$&    128,39	 & 83,46 &   & 98,43	 & 146,23 &   & 20,45	 & 11,06 &   & 2,75	 & 2,48 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

Table \ref{table:average_AS_GPU} shows the speedup obtained when solving a series of problems on the GPU using an adaptive scheme compared to a similar launch on CPU using 4 threads. Table \ref{table:average_MMAS_GPU} shows the speedup of the algorithm combining multiple evolvents and an adaptive scheme when solving a series of problems on GPU, compared to an adaptive scheme on CPU using 4 threads. Two evolvents and, accordingly, two processes were used; each process used $p$ threads, and calculations were performed on a single cluster node.

\begin{table}
                \caption{Speedup on GPU}
                \label{table:GKLS6_average_MMAS_GPU}
                \center
                \begin{tabular}{lcc}
                               %\hline\noalign{\smallskip}
                &  Iteration speedup &   Time speedup\\

                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,32	 & 6,83	  \\
                               $p=256$ &    38,96	 & 31,52  \\
							   $p=1024$&    128,39	 & 83,46  \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

%Следующий эксперимент будет проведен на 6-ти мерных задачах GKLS. В таблице \ref{table:GKLS6_average_MMAS_GPU} приведено ускорение по времени и по итерациям полученное при решении серии задач на GPU с использованием алгоритма объединяющего множественные развертки и адаптивную схему по сравнению с адаптивной схемой на CPU использующей 4 потока. Эксперимент проводился на трех узлах кластера и спользовалось до 6144 нитей.

@@

In this paper, we consider a parallel algorithm for solving multidimensional multiextremal optimization problems and its implementation on the GPU. In order to experimentally confirm the theoretical properties of the parallel algorithm under consideration, computational experiments were carried out on a series of several hundred test problems of different dimensions. The parallel algorithm demonstrates good acceleration, both in the GPU and CPU versions.

>>>>

Problems of multidimensional multiextremal optimization and numerical methods for their solution are considered. The general assumption is made about the function being optimized: it satisfies the Lipschitz condition with an a priori unknown constant. Many approaches to solving problems of this class are based on reducing the dimension of the problem; i.e. addressing a multidimensional problem by solving a family of problems with lower dimension. In this work, an adaptive dimensionality reduction scheme is investigated, and its implementation using graphic accelerators is proposed. Numerical experiments on several hundred test problems were carried out, and they confirmed acceleration in the developed GPU version of the algorithm.

@@

However, the potential for graphics accelerators to solve global optimization problems has not yet been fully realized. Using  GPUs, they basically parallelize nature-inspired optimization algorithms, which are somehow based on the idea of random search (see, for example, \cite{Ferreiro2013,Garcia2014,Langdon2011}). By virtue of their stochastic nature, algorithms of this type guarantee convergence to the global minimum only in the sense of probability, which differentiates them unfavorably from deterministic methods.

With regard to many deterministic algorithms of Lipschitzian global optimization with guaranteed convergence, parallel variants have been proposed \cite{Evtushenko2009,He2008,Paulavicius2011}. However, these versions of algorithms are parallelized on CPU using shared and/or distributed memory; presently, no GPU implementations have been made. For example, \cite{Paulavicius2011} describes parallelization of an algorithm based on the ideas of the branch and boundary method using MPI and OpenMP.

Within the framework of this research, we consider the problems of capturing the optimum, which are characterized by a lengthy period for calculating the values of objective function in comparison with the time needed for processing them. For example, objective function can be specified using systems of linear algebraic equations, systems of ordinary differential equations, etc. Currently, graphics accelerators can be used to solve problems of this type. Moreover, an accelerator can solve several such problems at once \cite{Kindratenko2014}; i.e., using GPU, one can calculate multiple function values simultaneously.

Thus, calculating the optimization criterion can be implemented on GPU, and the role of the optimization algorithm (running on CPU) consists in the effective selection of points for conducting parallel trials. This scheme of working with the accelerator is fully consistent with the work of the parallel global search algorithm developed at the Lobachevsky State University of Nizhni Novgorod and presented in a series of publications \cite{Barkalov2014,Barkalov2020,Barkalov2019,Grishagin2016,Grishagin2015,Grishagin2018,globalizerSystem}.

@@

In relation to global optimization methods, an operation that can be efficiently implemented on GPU is the parallel calculation of many values of the objective function at once. Naturally, this requires implementing a procedure for calculating the value of a function on GPU. Data transfers from CPU to GPU will be minimal: one need only transfer the coordinates of the trial points to GPU, and get back the function values at these points. Functions that determine the processing of trial results in accordance with the algorithm, and require work with large amounts of accumulated search information, can be effectively implemented on CPU.

@@

\ref{fig:Parallel_Computing} shows the general scheme of organizing calculations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\varphi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using CPU only. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

@@

The numerical experiments were carried out using the Lomonosov supercomputer (Lomonosov Moscow State University). Each supercomputer node included two quad-core processors Intel Xeon X5570, two NVIDIA Tesla X2070 and 12 Gb RAM. When building the Globalizer system for running on the Lomonosov supercomputer, the GCC 4.3.0 compiler, CUDA 6.5 and Intel MPI 2017 were used.

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time span for calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems \cite{Modorskii2016}, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series).

@@

Let us now evaluate the acceleration achieved using parallel GSA using an adaptive dimensionality reduction scheme based on the number $p$ of cores used.
Table \ref{table:average_MMAS_CPU} shows the speedup of the algorithm that combines multiple evolvents and an adaptive scheme for solving a series of problems on CPU, compared to the sequential launch of the GSA method. Two evolvents and, accordingly, two processes were used; each process used $p$ threads, and calculations were performed on a single cluster node.
% Table \ref{table:average_AS_CPU} shows the speedup obtained when solving a series of problems on the CPU, compared to GSA.

@@

\begin{table}
                \caption{Speedup on one GPU}
                \label{table:average_AS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    5,06  & 3,93  &   & 1,19 	 & 5,41  &   & 2,2	    & 1,92	 &   & 0,47	& 2,21 \\
                               $p=256$ &    19,91 & 14,95 &   & 11,9 	 & 23,36 &   & 8,28	& 6,91	     &   & 3,4	    & 6,1 \\
							   $p=1024$&    15,62 & 52,91 &   & 22,12	 & 55,22 &   & 5,21	& 19,95      &   & 2,18	& 4,82 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on two GPUs}
                \label{table:average_MMAS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,32	 & 6,83	 &   & 8,19	     & 10,66  &   & 2,44	 & 1,63	 &   & 1,33	 & 1,58 \\
                               $p=256$ &    38,96	 & 31,52 &   & 25,43	 & 48,23  &   & 8,01	 & 7,04	 &   & 2,31	 & 3,42 \\
							   $p=1024$&    128,39	 & 83,46 &   & 98,43	 & 146,23 &   & 20,45	 & 11,06 &   & 2,75	 & 2,48 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

Table \ref{table:average_AS_GPU} shows the speedup obtained when solving a series of problems on one GPU using an adaptive scheme compared to a similar launch on CPU using 4 threads. Table \ref{table:average_MMAS_GPU} shows the speedup of the algorithm combining multiple evolvents and an adaptive scheme when solving a series of problems on two GPUs compared to an adaptive scheme on CPU using 4 threads. Two evolvents and, accordingly, two processes were used; each process used $p$ threads on each GPU; all computations were performed on a single cluster node.

\begin{table}
                \caption{Speedup on six GPUs}
                \label{table:GKLS6_average_MMAS_GPU}
                \center
                \begin{tabular}{lccc}
                               %\hline\noalign{\smallskip}
                &  Iteration speedup & &   Time speedup \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,32	 & & 6,83	  \\
                               $p=256$ &    38,96	 & & 31,52  \\
															 $p=1024$&    128,39	 & & 83,46  \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

%Table \ref{table:average_AS_GPU} shows the speedup obtained when solving a series of problems on one GPU using an adaptive scheme compared to a similar launch on CPU using 4 threads. Table \ref{table:average_MMAS_GPU} shows the speedup of the algorithm combining multiple evolvents and an adaptive scheme when solving a series of problems on two GPUs compared to an adaptive scheme on CPU using 4 threads. Two evolvents and, accordingly, two processes were used; each process used $p$ threads on each GPU; all computations were performed on a single cluster node.

The last series of experiments has been carried out on 100 six-dimensional problems from the GKLS Simple class. 
Table \ref{table:GKLS6_average_MMAS_GPU} shows the speedup of the algorithm combining multiple evolvents and the adaptive scheme when solving the problems on 3 cluster nodes (using 2 GPUs per node, 6144 GPU threads in all) compared to the adaptive scheme on CPU using 4 threads.

%Заключительный эксперимент был проведен на 6-ти мерных задачах класса GKLS Simple. В таблице \ref{table:GKLS6_average_MMAS_GPU} приведено ускорение по времени и по итерациям полученное при решении серии задач на GPU с использованием алгоритма объединяющего множественные развертки и адаптивную схему по сравнению с адаптивной схемой на CPU использующей 4 потока. Эксперимент проводился на трех узлах кластера, шесть разверток = шесть проецессов, каждый процесс - p нитей. Максимальное число нитей 6144.

@@

In this paper, we consider a parallel algorithm for solving multidimensional multiextremal optimization problems and its implementation on GPU. In order to experimentally confirm the theoretical properties of the parallel algorithm under consideration, computational experiments were carried out on a series of several hundred test problems of different dimensions. The parallel algorithm demonstrates good speedup, both in the GPU and CPU versions.
